---
title: "PCA using SNPrelate package"
output: html_document
date: "2022-10-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I am following instructions from [this tutorial](https://www.bioconductor.org/packages/devel/bioc/vignettes/SNPRelate/inst/doc/SNPRelate.html). 

## Load libraries and data 

```{r}
library(snpStats)
library(tidyverse)
library(gdsfmt)
library(SNPRelate)
```

Load the data and convert to gds format. 

```{r}
bed.fn.m <- "../../Desktop/GWAStutorial/public/Genomics/108Malay_2527458snps.bed"
fam.fn.m <- "../../Desktop/GWAStutorial/public/Genomics/108Malay_2527458snps.fam"
bim.fn.m <- "../../Desktop/GWAStutorial/public/Genomics/108Malay_2527458snps.bim"

snpgdsBED2GDS(bed.fn.m, fam.fn.m, bim.fn.m, "test.gds")

bed.fn.i <- "../../Desktop/GWAStutorial/public/Genomics/105Indian_2527458snps.bed"
fam.fn.i <- "../../Desktop/GWAStutorial/public/Genomics/105Indian_2527458snps.fam"
bim.fn.i <- "../../Desktop/GWAStutorial/public/Genomics/105Indian_2527458snps.bim"

snpgdsBED2GDS(bed.fn.i, fam.fn.i, bim.fn.i, "test2.gds")

bed.fn.c <- "../../Desktop/GWAStutorial/public/Genomics/110Chinese_2527458snps.bed"
fam.fn.c <- "../../Desktop/GWAStutorial/public/Genomics/110Chinese_2527458snps.fam"
bim.fn.c <- "../../Desktop/GWAStutorial/public/Genomics/110Chinese_2527458snps.bim"

snpgdsBED2GDS(bed.fn.c, fam.fn.c, bim.fn.c, "test3.gds")

snpgdsSummary("test.gds")
```

Combine three population files into one. 
```{r}
fn1 <- "test.gds"
fn2 <- "test2.gds"
fn3 <- "test3.gds"
snpgdsCombineGeno(c(fn1, fn2, fn3), "test4.gds")
```

Get a summary of the file combined file and open it. 
```{r}
snpgdsSummary("test4.gds")
genofile <- snpgdsOpen("test4.gds")
```

## LD-based SNP pruning

It is suggested to use a pruned set of SNPs which are in approximate linkage equilibrium with each other to avoid the strong influence of SNP clusters in principal component analysis and relatedness analysis.

```{r}
set.seed(1000)

# Try different LD thresholds for sensitivity analysis
snpset <- snpgdsLDpruning(genofile, ld.threshold=0.2)

str(snpset)
names(snpset)

# Get all selected snp id
snpset.id <- unlist(unname(snpset))
head(snpset.id)
```

## PCA 

```{r}
# Run PCA
pca <- snpgdsPCA(genofile, snp.id=snpset.id, num.thread=2)
```

Calculate the percent of variation is accounted for by the top principal components.

```{r}
# variance proportion (%)
pc.percent <- pca$varprop*100
head(round(pc.percent, 2))
```

### No prior population information

```{r}
# make a data.frame
tab <- data.frame(sample.id = pca$sample.id,
    EV1 = pca$eigenvect[,1],    # the first eigenvector
    EV2 = pca$eigenvect[,2],    # the second eigenvector
    stringsAsFactors = FALSE)
head(tab)
```

```{r}
# Draw
plot(tab$EV2, tab$EV1, xlab="eigenvector 2", ylab="eigenvector 1")
```

### Prior population information

```{r}
sample.id <- read.gdsn(index.gdsn(genofile, "sample.id"))
pop_code <- read.gdsn(index.gdsn(genofile, "sample.annot/pop.group")) ## doesn't work
```

